---
title: "CS169_Final_Hammond"
author: "Evan Gray, Emilia Nathan, Tomer Wenderow"
date: "2024-12-03"
output: html_document
---

---
title: "Final_Project"
author: "Evan Gray"
date: "2024-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(ape)
library(qs)
library(tidyverse)
library(Seurat) #v4.1.0
library(scCustomize) #v0.7.0.9938
library(qs)
library(data.table)
library(mclust)
```


```{r}
library(qs)
dt = qs::qread('Hammond_et-al-2019_Seurat_Converted_v4.qs') 
counts = dt@assays$RNA@counts 
cell_meta = dt@meta.data
```

Whole dataset
```{r}
dt = qs::qread('Hammond_et-al-2019_Seurat_Converted_v4.qs') 
counts = dt@assays$RNA@counts 
cell_meta = dt@meta.data
hammond_seurat <- CreateSeuratObject(counts = counts, meta.data = cell_meta)
# Normalize the data
hammond_seurat <- NormalizeData(hammond_seurat)

# Identify variable features
hammond_seurat <- FindVariableFeatures(hammond_seurat, selection.method = "vst", nfeatures = 2000)

# Scale the data
hammond_seurat <- ScaleData(hammond_seurat)

# Perform PCA
hammond_seurat <- RunPCA(hammond_seurat, features = VariableFeatures(hammond_seurat))
#seurat_subset <- RunICA(seurat_subset, nics = 50)
```

```{r}
# Find neighbors
hammond_seurat <- FindNeighbors(hammond_seurat, dims = 1:10)

# Perform clustering
hammond_seurat <- FindClusters(hammond_seurat, resolution = 0.5)

# Visualize clusters using UMAP
hammond_seurat <- RunUMAP(hammond_seurat, dims = 1:10)
DimPlot(hammond_seurat, reduction = "umap", label = TRUE)
```

PCA embeddings of the hammond data
```{r}
hammond_embeddings <- hammond_seurat@reductions[["pca"]]@cell.embeddings
```

```{r}
wcss <- numeric()

for (k in 1:15) {
  kmeans_model <- kmeans(hammond_embeddings, centers = k, nstart = 10, iter.max = 10000)
  wcss[k] <- kmeans_model$tot.withinss
}

```

```{r}
elbow_plot <- data.frame(
  Clusters = 1:15,
  WCSS = wcss
)

ggplot(elbow_plot, aes(x = Clusters, y = WCSS)) +
  geom_line() +
  geom_point() +
  ggtitle("Elbow Method for Optimal Clusters (Hammond Data)") +
  xlab("Number of Clusters (k)") +
  ylab("Within-Cluster Sum of Squares")
```

Hammond K-Means UMAP
```{r}
hammond_clusters <- kmeans(hammond_embeddings, centers = 11, iter.max = 10000)
kmeans_clusters <- hammond_clusters$cluster
hammond_seurat$seurat_clusters <- as.factor(kmeans_clusters)
Idents(hammond_seurat) <- hammond_seurat$seurat_clusters
DimPlot(hammond_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of Hammond Clusters (K-Means)")
```

Average expression hammond k-means
```{r}
average_expression_kmeans <- AverageExpression(hammond_seurat, return.seurat = FALSE)
avg_expr <- average_expression_kmeans$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression value in descending order
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  top_genes <- head(sorted_genes, 5)
  
  # Dataframe with results
  data.frame(Gene = names(top_genes), Expression = top_genes)
})

# Combine results
most_expressed_genes_df_hk <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))

most_expressed_genes_df_hk
```

Save the hammond k-means data
```{r}
write.csv(most_expressed_genes_df_hk, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/Hammond_K-Means_Gene_Clusters.csv", row.names = FALSE)
```

Hammond PCA explained
```{r}
pca_stdev <- hammond_seurat@reductions[["pca"]]@stdev
# Variance explained by each Principal Component
variance_explained <- (pca_stdev^2) / sum(pca_stdev^2) * 100
cumulative_variance <- cumsum(variance_explained)

# Plot cumulative variance
ggplot(data.frame(PC = 1:length(cumulative_variance), CumulativeVariance = cumulative_variance),
       aes(x = PC, y = CumulativeVariance)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Cumulative Variance Explained by Principal Components (Hammond)",
       x = "Principal Component",
       y = "Cumulative Percentage of Variance Explained")

pc_variance_df <- data.frame(
  PC = 1:length(variance_explained),
  Variance = variance_explained
)

ggplot(pc_variance_df, aes(x = PC, y = Variance)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Variance Explained by Principal Components",
       x = "Principal Component",
       y = "Percentage of Variance Explained")

total_var <- 0

for (i in 1:50) {
  cat(sprintf("PC%d explains %.2f%% of the variance\n", i, variance_explained[i]))
  total_var <- total_var + variance_explained[i]
}
total_var

```

Hammond Hierarchical Clustering
```{r}
hammond_seurat <- FindClusters(hammond_seurat, method = "hierarchical", algorithm = 3)
```

Save Hierarchical Clusters
```{r}
hammond_hierarchical_clusters <- as.factor(hammond_seurat@meta.data[["seurat_clusters"]])
```

```{r}
#hammond_seurat <- RunUMAP(hammond_seurat, dims = 1:10)

# Visualize clusters
DimPlot(hammond_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of Hammond Clusters (Hierarchical)")
```

Visualize Hierarchical Clusters
```{r}
DimPlot(hammond_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of Hierarchical Clusters (Hammond)")
```

```{r}
hammond_seurat@meta.data[["seurat_clusters"]]
```

Dendrogram
```{r}
# Build the hierarchical dendrogram
hammond_seurat <- BuildClusterTree(hammond_seurat, dims = 1:10)
# Plot the dendrogram
PlotClusterTree(hammond_seurat)
```

Hammond average expressed genes per cluster
```{r}
average_expression <- AverageExpression(hammond_seurat, return.seurat = FALSE)
avg_expr <- average_expression$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression value in descending order
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  
  # Get the top 5 genes and their expression values
  top_genes <- head(sorted_genes, 5)
  
  # Create a data frame with the results
  data.frame(Gene = names(top_genes), Expression = top_genes)
})

# Combine the results into a single data frame
most_expressed_genes_df_hh <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))

# View the results
most_expressed_genes_df_hh
```

Save hierarchical clusters
```{r}
write.csv(most_expressed_genes_df_hh, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/Hammond_Hierarchical_Gene_Clusters.csv", row.names = FALSE)
```

Hammond GMMs
```{r}
# Took a long time to run and needed to do top 35 PCs
#pca_data <- hammond_seurat@reductions$pca@cell.embeddings[, 1:35]
#gmm_result <- Mclust(pca_data) 
#saveRDS(gmm_result, "/Users/evangray/Library/CloudStorage/Box-Box/hammond_gmm_result.rds")
```

Save GMM so do not need to run again to analyze results
```{r}
hammond_gmm <- readRDS("/Users/evangray/Library/CloudStorage/Box-Box/hammond_gmm_result.rds")
```

For visualization
```{r}
hammond_seurat$seurat_clusters <- as.factor(hammond_gmm[["classification"]])
Idents(hammond_seurat) <- hammond_seurat$seurat_clusters
```

Average Expression Hammond GMM
```{r}
average_expression_gmm <- AverageExpression(hammond_seurat, return.seurat = FALSE)
avg_expr <- average_expression_gmm$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression value in descending order
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  # Top 5 genes
  top_genes <- head(sorted_genes, 5)
  
  # Create a data frame with the results
  data.frame(Gene = names(top_genes), Expression = top_genes)
})
most_expressed_genes_df_hgmm <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))

most_expressed_genes_df_hgmm
```

save hammond gmm gene clusters
```{r}
write.csv(most_expressed_genes_df_hgmm, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/Hammond_GMM_Gene_Clusters.csv", row.names = FALSE)
```

Prepare results
```{r}
hammond_seurat$seurat_clusters <- as.factor(hammond_gmm[["classification"]])
Idents(hammond_seurat) <- hammond_seurat$seurat_clusters
```

Plot UMAP (GMM)
```{r}
# Visualize clusters
DimPlot(hammond_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of Hammond Clusters (GMM)")
```


Paper clustering results
```{r}
hammond_seurat$seurat_clusters <- as.factor(hammond_seurat@meta.data[["Paper_Cluster"]])
Idents(hammond_seurat) <- hammond_seurat$seurat_clusters
DimPlot(hammond_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of Hammond Clusters (Paper Results)")
```

Paper Top Genes
```{r}
average_expression_paper <- AverageExpression(hammond_seurat, return.seurat = FALSE)
avg_expr <- average_expression_paper$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression value in descending order
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  # Top 5 genes
  top_genes <- head(sorted_genes, 5)
  data.frame(Gene = names(top_genes), Expression = top_genes)
})

# Combine the results into a single data frame
most_expressed_genes_df_paper <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))
most_expressed_genes_df_paper
```

save hammond paper gene clusters
```{r}
write.csv(most_expressed_genes_df_paper, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/Hammond_Paper_Gene_Clusters.csv", row.names = FALSE)
```

Hammond GMM clusters
```{r}
# Get the uncertainty and assignments
posterior_probs <- hammond_gmm$z
cluster_assignments <- hammond_gmm$classification
uncertainty <- 1 - apply(posterior_probs, 1, max)

# Cumulative uncertainty per cluster
total_uncertainty_per_cluster <- tapply(uncertainty, cluster_assignments, sum)
total_uncertainty_per_cluster
```

Hammond UMAP by Age
```{r}
DimPlot(hammond_seurat, reduction = "umap", group.by = "Age", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of Hammond Clusters (Age)")
```
