---
title: "CS169_Final_GLP"
author: "Evan Gray, Emilia Nathan, Tomer Wenderow"
date: "2024-12-01"
output: html_document
---

```{r}
library(ggplot2)
library(ape)
library(qs)
library(tidyverse)
library(Seurat) #v4.1.0
library(scCustomize) #v0.7.0.9938
library(qs)
library(data.table)
library(mclust)
```

Create the Seurat Object
```{r}
#meta_path <- "/Users/evangray/Library/CloudStorage/Box-Box/metadata_mg_filtered.csv"
#expression_path <- "/Users/evangray/Library/CloudStorage/Box-Box/expression_mg_filtered.csv"

#expression_data <- read.csv(expression_path, row.names = 1)
#metadata <- read.csv(meta_path, header = TRUE, row.names = 1)
#metadata<- metadata[-1, ]
```

```{r}
#glp_seurat <- CreateSeuratObject(counts = expression_data, meta.data = metadata)
#saveRDS(glp_seurat, file = "/Users/evangray/Library/CloudStorage/Box-Box/glp_seurat_full.rds")
```

Read in the data
```{r}
glp_seurat <- readRDS("/Users/evangray/Library/CloudStorage/Box-Box/glp_seurat_full.rds")
```

Preprocessing and dimensionality reduction
```{r}
# Normalize the data
glp_seurat <- NormalizeData(glp_seurat)

# Identify variable features
glp_seurat <- FindVariableFeatures(glp_seurat, selection.method = "vst", nfeatures = 2000)

# Scale the data
glp_seurat <- ScaleData(glp_seurat)

# Perform PCA
glp_seurat <- RunPCA(glp_seurat, features = VariableFeatures(glp_seurat))
glp_seurat <- FindNeighbors(glp_seurat, dims = 1:10)
```

PCA explained
```{r}
pca_stdev <- glp_seurat@reductions[["pca"]]@stdev
# Variance explained by each PC
variance_explained <- (pca_stdev^2) / sum(pca_stdev^2) * 100
cumulative_variance <- cumsum(variance_explained)

# Plot cumulative variance
ggplot(data.frame(PC = 1:length(cumulative_variance), CumulativeVariance = cumulative_variance),
       aes(x = PC, y = CumulativeVariance)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Cumulative Variance Explained by PCs",
       x = "Principal Component",
       y = "Cumulative Percentage of Variance Explained")
pc_variance_df <- data.frame(
  PC = 1:length(variance_explained),
  Variance = variance_explained
)

# Plot the variance explained
library(ggplot2)
ggplot(pc_variance_df, aes(x = PC, y = Variance)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Variance Explained by Principal Components",
       x = "Principal Component",
       y = "Percentage of Variance Explained")

total_var <- 0
# Print variance explained by first 10 PCs
for (i in 1:50) {
  cat(sprintf("PC%d explains %.2f%% of the variance\n", i, variance_explained[i]))
  total_var <- total_var + variance_explained[i]
}
total_var
```

GLP K-Means
```{r}
glp_embeddings <- glp_seurat@reductions[["pca"]]@cell.embeddings
```

Elbow plot
```{r}
# Build an elbow plot to determine optimal clusters
wcss <- numeric()

for (k in 1:15) { 
  kmeans_model <- kmeans(glp_embeddings, centers = k, nstart = 10, iter.max = 10000)
  wcss[k] <- kmeans_model$tot.withinss
}

elbow_plot <- data.frame(
  Clusters = 1:15,
  WCSS = wcss
)

ggplot(elbow_plot, aes(x = Clusters, y = WCSS)) +
  geom_line() +
  geom_point() +
  ggtitle("Elbow Method for Optimal Clusters (GLP Data)") +
  xlab("Number of Clusters (k)") +
  ylab("Within-Cluster Sum of Squares")
```

Final K-Means Clustering
```{r}
glp_clusters <- kmeans(glp_embeddings, centers = 11, iter.max = 10000)
glp_kmeans_clusters <- glp_clusters$cluster
glp_seurat$seurat_clusters <- as.factor(glp_kmeans_clusters)
Idents(glp_seurat) <- glp_seurat$seurat_clusters
```

Top 5 Expressed Genes from each cluster - K-Means
```{r}
average_expression_glp_km <- AverageExpression(glp_seurat, return.seurat = FALSE)
avg_expr <- average_expression_glp_km$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  # Top 5 Genes
  top_genes <- head(sorted_genes, 5)
  data.frame(Gene = names(top_genes), Expression = top_genes)
})

# Combine results
most_expressed_genes_df_gk <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))


most_expressed_genes_df_gk
```
save glp gene clusters - K-Means
```{r}
write.csv(most_expressed_genes_df_gk, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/GLP_K-Means_Gene_Clusters.csv", row.names = FALSE)
```

Plot UMAP (K-Means)
```{r}
glp_clusters <- kmeans(glp_embeddings, centers = 11, iter.max = 10000)
glp_kmeans_clusters <- glp_clusters$cluster
glp_seurat$seurat_clusters <- as.factor(glp_kmeans_clusters)
Idents(glp_seurat) <- glp_seurat$seurat_clusters
# Visualize clusters
DimPlot(glp_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of GLP Clusters (K-Means)")
```


GLP Hierarchical Clustering
```{r}
glp_seurat <- FindClusters(glp_seurat, method = "hierarchical", algorithm = 3)
```
Save Hierarchical Clusters
```{r}
glp_hierarchical_clusters <- as.factor(glp_seurat@meta.data[["seurat_clusters"]])
```


GLP Dendrogram
```{r}
# Build the hierarchical dendrogram
glp_seurat <- BuildClusterTree(glp_seurat, dims = 1:10)

# Plot the dendrogram
PlotClusterTree(glp_seurat)
```

Top 5 Expressed Genes from each cluster - Hierarchical
```{r}
average_expression_glp_h <- AverageExpression(glp_seurat, return.seurat = FALSE)
avg_expr <- average_expression_glp_h$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression value
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  # Top 5 genes
  top_genes <- head(sorted_genes, 5)
  data.frame(Gene = names(top_genes), Expression = top_genes)
})
# Get them for each cluster
most_expressed_genes_df_gh <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))


most_expressed_genes_df_gh
```

save glp gene clusters - Hierarchical
```{r}
write.csv(most_expressed_genes_df_gh, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/GLP_Hierarchical_Gene_Clusters.csv", row.names = FALSE)
```

Plot UMAP (Hierarchical)
```{r}
glp_seurat$seurat_clusters <- as.factor(glp_hierarchical_clusters)
Idents(glp_seurat) <- glp_seurat$seurat_clusters
DimPlot(glp_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of GLP Clusters (Hierarchical)")
```

Do the pruning
```{r}
custom_clusters <- c(
  "1" = "c1", "12" = "c1", "6" = "c1", "2" = "c2",
  "7" = "c2", "3" = "c2", "5" = "c10", "8" = "c2", "4" = "c2", "10" = "c2",
  "11" = "c3",
  "17" = "c4",
  "13" = "c5",
  "15" = "c6",
  "9" = "c7",
  "14" = "c9",
  "16" = "c8"
)
custom_cs <- custom_clusters[glp_seurat$seurat_clusters]
names(custom_cs) <- names(glp_seurat$seurat_clusters)
glp_seurat$custom_clusters <- custom_cs
```

Plot UMAP (Hierarchical Pruned Clusters)
```{r}
glp_seurat$seurat_clusters <- as.factor(glp_hierarchical_clusters)
Idents(glp_seurat) <- glp_seurat$seurat_clusters
DimPlot(glp_seurat, reduction = "umap", group.by = "custom_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of GLP Clusters (Hierarchical with Pruning)")
```

Top 5 Expressed Genes from each cluster - GMM
```{r}
glp_seurat$seurat_clusters <- as.factor(glp_seurat$custom_clusters)
Idents(glp_seurat) <- glp_seurat$seurat_clusters

average_expression_glp_ghp <- AverageExpression(glp_seurat, return.seurat = FALSE)
avg_expr <- average_expression_glp_ghp$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  # Top 5 genes
  top_genes <- head(sorted_genes, 5)
  data.frame(Gene = names(top_genes), Expression = top_genes)
})

# Get them for each cluster
most_expressed_genes_df_ghp <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))

most_expressed_genes_df_ghp
```

save glp gene clusters - Hierarchical with Pruning
```{r}
write.csv(most_expressed_genes_df_ghp, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/GLP_Hierarchical_Pruning_Gene_Clusters.csv", row.names = FALSE)
```

Run GMM Clustering
```{r}
# Took a long time to run
#pca_data_glp <- glp_seurat@reductions$pca@cell.embeddings[, 1:35]
#gmm_result_glp <- Mclust(pca_data_glp) 
#saveRDS(gmm_result_glp, "/Users/evangray/Library/CloudStorage/Box-Box/glp_gmm_result.rds")
```

Load glp GMM
```{r}
glp_gmm <- readRDS("/Users/evangray/Library/CloudStorage/Box-Box/glp_gmm_result.rds")
```

Prepare results
```{r}
glp_seurat$seurat_clusters <- as.factor(glp_gmm[["classification"]])
Idents(glp_seurat) <- glp_seurat$seurat_clusters
```

Top 5 Expressed Genes from each cluster - GMM
```{r}
average_expression_glp_gmm <- AverageExpression(glp_seurat, return.seurat = FALSE)
avg_expr <- average_expression_glp_gmm$RNA
most_expressed_genes <- apply(avg_expr, 2, function(cluster_expr) {
  # Sort the genes by expression
  sorted_genes <- sort(cluster_expr, decreasing = TRUE)
  # Top 5 genes
  top_genes <- head(sorted_genes, 5)
  data.frame(Gene = names(top_genes), Expression = top_genes)
})

# Ge
most_expressed_genes_df_ggmm <- do.call(rbind, lapply(names(most_expressed_genes), function(cluster) {
  cbind(Cluster = cluster, most_expressed_genes[[cluster]])
}))
most_expressed_genes_df_ggmm
```
save glp gene clusters - GMM
```{r}
write.csv(most_expressed_genes_df_ggmm, file = "/Users/evangray/Library/CloudStorage/Box-Box/bioinformatics_final/GLP_GMM_Gene_Clusters.csv", row.names = FALSE)
```

Uncertainty totals - GMM
```{r}
# Get the uncertainty and assignments
posterior_probs <- glp_gmm$z
cluster_assignments <- glp_gmm$classification
uncertainty <- 1 - apply(posterior_probs, 1, max)

# Cumulative uncertainty per cluster
total_uncertainty_per_cluster <- tapply(uncertainty, cluster_assignments, sum)
total_uncertainty_per_cluster
```


Plot UMAP (GMM)
```{r}
glp_seurat$seurat_clusters <- as.factor(glp_gmm[["classification"]])
Idents(glp_seurat) <- glp_seurat$seurat_clusters
# Visualize clusters
DimPlot(glp_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of GLP Clusters (GMM)")
```


```{r}
# Visualize clusters
DimPlot(glp_seurat, reduction = "umap", group.by = "biosample_id", label = TRUE, repel = TRUE) +
  ggtitle("UMAP Visualization of GLP Clusters (Age)")
```
